The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-03-21 ms.Rmd
restructuring intro, adding figure referencing

##### Past work and other software - the USP of ZOON and requirements

Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt or R as the first choice software for their analyses. 
The software market is much diverse, however, including a variety of software developed specifically for SDM (e.g. MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()). 
Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not. 
Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science. 
There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community. 
Perhaps this is because of the unfamiliarity or assumed complexity of these tools. 
The SDM package [@Naimi2016] has some functionality for users to incorporate new modelling methods into their analyses.

<!-- The Solution: ZOOOOOON!  -->
The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
Each of these steps is carried out by one or more community-contributed software modules. 
Firstly, occurrence data must be read into the software package or downloaded from online repositories.
Occurrence usually takes the form of presence-absence data or presence-only data, though abundance data is also used.
Secondly, predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
Thirdly, a wide variety of processes are often applied to the occurrence and covariate data.
These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
Fourthly, once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
Finally, model outputs are needed.
The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

<!-- How It Works -->
### Implementing a `zoon` workflow

Module types (and examples) in a diagram 
```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the module types. B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module'). C) Chaining and listing modules of the same type. D) An example workflow, corresponding to Example 1 below."}

@@

#### Re-running a fairly simple SDM, tweaking some different stuff and looking at the results

>>>>

Core, soluble problems:

<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->

Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt or R as the first choice software for their analyses. 
The software market is much diverse, however, including a variety of software developed specifically for SDM (e.g. MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()).

Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not. 
Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.

The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community.
Perhaps this is because of the unfamiliarity or assumed complexity of these tools.

The SDM package [@Naimi2016] implements a number of modern SDM approaches in a simple modular framework.
This package also has some functionality for users to incorporate new modelling methods into their analyses.
There is, however still a signifiact technical barrier to those wishing to incorporate their methods, and to share them more widely.

#### The ZOON project

The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
First, we describe the modular structure of zoon workflows, illustrate how to download and upload software modules, and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and illustrate the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

<!-- The Solution: ZOOOOOON!  -->
### The `zoon` R package

Versions 0.4-22 and later of the `zoon` R package can be installed directly from CRAN.

#### Constructing a `zoon` workflow

The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
Each of these steps is carried out by one or more community-contributed software modules.

Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow.

```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the module types. B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module'). C) Chaining and listing modules of the same type. D) An example workflow, corresponding to Example 1 below.\\label{fig:workflows}"}

Firstly, occurrence data must be read into the software package or downloaded from online repositories.
Occurrence usually takes the form of presence-absence data or presence-only data, though abundance data is also used.
Secondly, predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
Thirdly, a wide variety of processes are often applied to the occurrence and covariate data.
These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
Fourthly, once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
Finally, model outputs are needed.
The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

@@

#### Example 1. A fairly simple SDM, tweaking some different stuff and looking at the results
